Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-02-21 RNARedPrint-Bioinfo2018.tex
Added complexity improvement Moved #P hardness result

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
  \let\itemize\compactitem
  \let\enditemize\endcompactitem
  \let\enumerate\compactenum
  \let\endenumerate\endcompactenum
  \let\description\compactdesc
  \let\enddescription\endcompactdesc
  \pltopsep=1pt
  \plitemsep=1pt
  \plparsep=1pt  
\usepackage{hyperref} 
\usepackage{xspace} 
\usepackage{amsmath,amssymb} 
\usepackage{bm}
\usepackage{verbatim}
\usepackage{longtable}
\usepackage[vlined]{algorithm2e}

\usepackage{stmaryrd}

@@

@@

%\newtheorem{algorithm}{Algorithm}
\newtheorem{axiom}{Axiom}
\newtheorem{hypothesis}{Working Hypothesis}

@@

\begin{figure*}[t]
{\centering\includegraphics[width=.8\textwidth]{Figs/Workflow}\\}
\caption{General outline of \ourprog{} for base pair-based energy models. From a set of target secondary structures (i), base-pairs are merged (ii) into a (base pair) dependency graph (iii) and transformed into a tree decomposition (iv). The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v). An adaptive scheme learns weights to achieve targeted energies and \GCb-content, leading to the production of suitable designs (vi).}
\label{fig:workflow}
\end{figure*}

@@

We consider a fixed set of target RNA structures
$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$. $\R$ induces a
\Def{base pair dependency graph} $G_{\R}$ with nodes $\{1,\dots,n\}$
and edges $S\bigcup_{\ell\in[1,k]} R_\ell$, which describe the minimal
dependencies present in all relevant settings due to the requirement
of canonical base pairing.

@@

The \Def{dependency graph} of the set $\F$ of functions (on
sequences of length $n$) is the \Def{hypergraph}
$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. 
Our algorithms will critically rely on a \Def{tree decomposition} of the dependency graph, which we define below.
\begin{definition}[Tree decomposition and width]
  \label{def:treedecomp}
  Given a hypergraph $G=(X, E)$, a \Def{tree decomposition} is a
  pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and
  $\chi(v)\subseteq X$ is the set of vertices from $G$ assigned to the node $v\in T$, such
  that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item  $\forall x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
\item $\forall e\in E$ of $G$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}
The \Def{width} of a tree decomposition $(T,\chi)$ is defined as
$\width(T,\chi) = \min_{u\in T} |\chi(u)| - 1 $. The \Def{treewidth}
of $G$ is the smallest width in any tree decomposition of $G$.
\end{definition}
\section{Counting valid designs is \#{\sf P}-hard}\label{sec:counting}

%%% reformulate later
%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given base pair dependency graph $G=(V,E)$. Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$. As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its base pair dependency graph is bipartite, which can be tested in linear time.
Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides. The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.

  Let us now hypothesize the existence of a polynomial-time algorithm $\mathcal{A}$ for \NumDesign over strongly-connected graphs $G$. Now consider the polynomial-time algorithm $\mathcal{A}'$ that first executes $\mathcal{A}$ on $G$ to produce $\NumDesign(G)$, and returns $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$. We conclude that $\mathcal{A}'$ solves $\#{\sf BIS}$ in polynomial-time. This means that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
\end{proof}

@@

For our algorithmic description, we translate the concepts of
Section~\ref{sec:problem-statement} to the formalism of constraint networks, here
specialized as RNA design network. This allows us to base our
algorithm on cluster tree elimination (CTE) by Rina
Dechter~\citep{Dechter2013}.
%
In the RNA design network, sequences $S\in\S_n$ are isomorphically
modeled as assignments $\val_S$ of bases to $n$ variables, each associated to
a sequence position. The functions $f\in\F$ of
Section~\ref{sec:problem-statement}, are interpreted as functions on the network's variables.

@@

     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\;
     $p \gets p\ \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
%
%     $p \gets 1$\;
%     \For{$f\in \phi(v)$}{
%  	      $p \gets p \cdot \exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$\; 
%     }
%     \For{$v\to{}w\in T$}{
%  	      $p \gets p \cdot \evalfor{\Message{v}{w}}{\substitute{\val}{\val'}}$\; 
%     }
  	  $r \gets r - p$\; 
  	  \If{$r<0$}{
  	    $a_{\rm res} \gets \substitute{\val}{\val'}$\;
  	  \For{$(v\to{}w)\in T$}{
  	      $a_{\rm res} \gets \substitute{a_{\rm res}}{\Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)}$\; 
     }
  	  \Return {$a_{\rm res}$}
  	  \;}
 }
 }
 \caption{Stochastic backtrack algorithm for assignments in the Boltzmann distribution.}\label{alg:sampling}
\end{algorithm}
%For example, for design to a single non-crossing target in the nearest neighbor energy model, the complexity of Algs.~\ref{alg:pf} and \ref{alg:sampling} degrades to linear time (as reported for IncaRNAtion), since there are only $O(n)$ many functions and nodes (in the sequence length $n$). Moreover, the dependencies are tree-like owing to the tree-like non-crossing structure; this implies a constantly bounded maximum treewidth.

As already noted by \citet{Flamm2001}, the
requirement of canonical base pairing induces a bi-partition of each connected
component of the base pair dependency graph, such that the variables of one
partition are restricted to values $\{\Ab,\Gb\}$, but for other
partition, to values $\{\Cb,\Ub\}$. This suggests a strong
optimization of our algorithms, which reduces the basis in the
exponential factor from 4 to 2. The price for this reduction is a
binary choice for each of $c$ connected component. For each of the
$2^c$ combinations, the partition functions can be computed separately
in $O(n\,m\, 2^{w+1})$ time. This strategy is applied only if
$c< w+1$. 
%% SW: This could be optimized further by applying the strategy only to (optimally) selected components
Finally, we compute the sums of the partition functions for all possible combinations.

\begin{theorem}[Complexities]
  For sequence length $n$, $k$ target structures, treewidth $w$ and
  a base pair dependency graph having $c$ connected components, $t$ sequences
  are sampled from the Boltzmann distribution  in
  %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
  $\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
\end{theorem}

@@

% To illustrate the expressivity of the framework, we outline the construction of function sets that express two different energy models for the energies $E(S;R_\ell)$ for encoding
% \begin{equation}
%   \label{eq:nussinov-network-energy}
%   \energy{\network}(S) = \sum_{\ell\in[1,k]} w_\ell E(S;R_\ell)
% \end{equation}
In simple \Def{base pair-based energy models}, energy is defined as
the sum of base pair (pseudo-)energies. If base pair energies
$\Ebp{k}{i,j,x,y}$ (where $i$ and $j$ are sequence positions, $x$ and
$y$ are bases in $\Sigma$) are given for each target structure $\ell$,
s.t.  $ E(S;R_\ell) := \sum_{(i,j)\in R_\ell} \Ebp{k}{i,j,S_i,S_j}$,
we encode the network energy by the set of functions $f$ for each base
pair $(i,j)\in R_\ell$ of each input structure $R_\ell$ that evaluate
to $\ln(\pi_\ell) \Ebp{k}{i,j,\val(x_i),\val(x_j)}$ under assignment $\val$;
here, $\pi_\ell>0$ is a weight that controls the
influence of structure $\ell$ on the sampling (as elaborated later).

@@

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, \emph{i.e.} such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp. Mat.~\ref{sec:weight_derivatives}). More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}. 
In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(e^\beta,\dots,e^\beta)$ and, at each iteration, generates a sample $\mathcal{S}$ of sequences. The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$. 
$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$. In practice, the constant $\gamma>1$ is chosen empirically to achieve effective optimization. 
While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior (choosing $\gamma=1.2$).

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rationale for a rejection approach follows from: 
i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}). 
%Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
For base pair-based energy models and special base pair dependency graph
(paths, cycles\ldots) this property rigorously follows from analytic
combinatorics, see \citet{Bender1983} and
\citet{Drmota1997}. In such cases, the expected number of
rejections before reaching the targeted energies remains constant when
$\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
$\varepsilon=0$. The \GCb-content of designs can also be controlled,
jointly and in a similar way, as done in
\Software{IncaRNAtion}~\citep{Reinharz2013}.

@@

Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps. At each step, we resampled (uniformly at random) the positions of a randomly selected component in the base pair dependency graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
%
Once again, for all instances, we observe a positive improvement of the quality of Boltzmann designs over uniform ones, suggesting a superior potential for optimization ($\sim 32\%$ avg improvement, Table~\ref{tab:benchmark-results}, optimized). This improvement is however more modest than in the absence of optimization, but we attribute this relative decay to the uniform nature of the resampling strategy, which somehow ends up canceling the benefits of the initial Boltzmann distribution, by uniformizing the sequence over longer runs. This suggests an alternative resampling strategy based on Boltzmann-distributed steps.

@@

@@

\newpage
\onecolumn

@@

In addition, we generated four sets of instances of increasing
complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
respectively specify 3,4,5, and 6 target structures for sequence
length 100.  For each instance (100 instances per set), we generated a
set of $k$ ($k=3,\dots,6$) compatible structures as follows
\begin{itemize}
\item Generate a random sequence of length 100;
\item Compute its minimum free energy structure (ViennaRNA package);
\item Add the new structure to the instances if the resulting base pair dependency graph is bipartite;
\item Repeat until $k$ structures are collected.
\end{itemize}
For each instance, we generated the dependency graphs in the base pair
model and in the stacking model. Then, we performed tree decomposition
(using strategy ``GreedyFillIn'' of LibTW~\citep{Dijk2006}) on each dependency
graph. The obtained treewidths are reported in
Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
corresponding run-times of the tree decomposition.
Fig.~\ref{fig:td-example} shows the tree decompositions for an example
instance from set RF3.

@@

>>>>

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
  \let\itemize\compactitem
  \let\enditemize\endcompactitem
  \let\enumerate\compactenum
  \let\endenumerate\endcompactenum
  \let\description\compactdesc
  \let\enddescription\endcompactdesc
  \pltopsep=1pt
  \plitemsep=1pt
  \plparsep=1pt  
\usepackage{hyperref} 
\usepackage{xspace} 
%\usepackage{geometry} 
\usepackage{amsmath,amssymb} 
\usepackage{bm}
\usepackage{verbatim}
\usepackage{longtable}
\usepackage[vlined]{algorithm2e}

\usepackage{xifthen}
\usepackage{stmaryrd}

@@

\hypersetup{
    bookmarks=false,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in Acrobat’s bookmarks
    pdftoolbar=true,        % show Acrobat’s toolbar?
    pdfmenubar=true,        % show Acrobat’s menu?
    pdffitwindow=true,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red!70!black,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue!50!black,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=green!50!black           % color of external links
}

@@

%\newtheorem{algorithm}{Algorithm}
\newtheorem{axiom}{Axiom}
\newtheorem{hypothesis}{Working Hypothesis}
\renewenvironment{proof}[1][]{\noindent \em Proof\ifthenelse{\equal{#1}{}}{}{ (#1)}:~}{}

@@

\begin{figure*}[t]
{\centering\includegraphics[width=.8\textwidth]{Figs/Workflow}\\}
\caption{General outline of \ourprog{} for base pair-based energy models. From a set of target secondary structures (i), base-pairs are merged (ii) into a compatibility graph (iii) and transformed into a tree decomposition (iv). The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v). An adaptive scheme learns weights to achieve targeted energies and \GCb-content, leading to the production of suitable designs (vi).}
\label{fig:workflow}
\end{figure*}

@@

We consider a fixed set of target RNA structures
$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$. $\R$ induces a
\Def{compatibility graph} $G_{\R}$ with nodes $\{1,\dots,n\}$ and
edges $S\bigcup_{\ell\in[1,k]} R_\ell$. While we typically consider
additional dependencies (in the dependency graph), the edges in the
compatibility graph pose constraints on the possible bases due to
canonical base pairing, which are present in all considered settings.

@@

The \Def{dependency graph} of the set $\F$ of functions (on
sequences of length $n$) is the \Def{hypergraph}
$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$. 
Our algorithms will critically rely on a \Def{tree decomposition} of the dependency graph, which we define below.
\begin{definition}[Tree decomposition and width]
  \label{def:treedecomp}
  Given a hypergraph $G=(X, E)$, a \Def{tree decomposition} is a
  pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and
  $\chi(v)\subseteq X$ is the set of vertices from $G$ assigned to the node $v\in T$, such
  that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item  $\forall x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
\item $\forall e\in E$ of $G$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}
The \Def{width} of a tree decomposition $(T,\chi)$ is defined as
$\width(T,\chi) = \min_{u\in T} |\chi(u)| - 1 $. The \Def{treewidth}
of $G$ is the smallest width in any tree decomposition of $G$.
\end{definition}

@@

For our algorithmic description, we translate the concepts of
Section~\ref{sec:problem-statement} to the formalism of constraint networks, here
specialized as RNA design network. This allows us to base our
algorithm on the cluster tree elimination (CTE) of~\citet{Dechter2013}.
%
In the RNA design network, sequences $S\in\S_n$ are isomorphically
modeled as assignments $\val_S$ of bases to $n$ variables, each associated to
a sequence position. The functions $f\in\F$ of
Section~\ref{sec:problem-statement}, are interpreted as functions on the network variables.

@@

     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\;
     $p \gets p\ \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
%
%     $p \gets 1$\;
%     \For{$f\in \phi(v)$}{
%  	      $p \gets p \cdot \exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$\; 
%     }
%     \For{$v\to{}w\in T$}{
%  	      $p \gets p \cdot \evalfor{\Message{v}{w}}{\substitute{\val}{\val'}}$\; 
%     }
  	  $r \gets r - p$\; 
  	  \If{$r<0$}{
  	    $a_{\rm res} \gets \substitute{\val}{\val'}$\;
  	  \For{$(v\to{}w)\in T$}{
  	      $a_{\rm res} \gets a_{\rm res} \mid \Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)$\; 
     }
  	  \Return {$a_{\rm res}$}
  	  \;}
 }
 }
 \caption{Stochastic backtrack algorithm for assignments in the Boltzmann distribution.}\label{alg:sampling}
\end{algorithm}
%For example, for design to a single non-crossing target in the nearest neighbor energy model, the complexity of Algs.~\ref{alg:pf} and \ref{alg:sampling} degrades to linear time (as reported for IncaRNAtion), since there are only $O(n)$ many functions and nodes (in the sequence length $n$). Moreover, the dependencies are tree-like owing to the tree-like non-crossing structure; this implies a constantly bounded maximum treewidth.

\begin{theorem}[Complexities]\label{th:complexities}
  For sequence length $n$, $k$ target structures, treewidth $w$ and
  a compatibility graph having $c$ connected components, $t$ sequences
  are generated from the Boltzmann distribution  in
  $O( n\, k \, 4^{w+1} + t\, n\, k )$ time.
  %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
  %$\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
\end{theorem}
As shown in Supp. Mat.~\ref{sec:improvedComplexity}, the complexity of the precomputation can be further improved to 
$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ is the maximum number of connected components represented in a node of the tree decomposition ($c\le w+1$).

@@

% To illustrate the expressivity of the framework, we outline the construction of function sets that express two different energy models for the energies $E(S;R_\ell)$ for encoding
% \begin{equation}
%   \label{eq:nussinov-network-energy}
%   \energy{\network}(S) = \sum_{\ell\in[1,k]} w_\ell E(S;R_\ell)
% \end{equation}
In simple \Def{base pair-based energy models}, energy is defined as the sum of base pair (pseudo)energies. If base pair energies $\Ebp{k}{i,j,x,y}$ (where
$i$ and $j$ are sequence positions, $x$ and $y$ are bases in $\Sigma$)
are given for each target structure $\ell$, s.t.
$ E(S;R_\ell) := \sum_{(i,j)\in R_\ell} \Ebp{k}{i,j,S_i,S_j}$, we
encode the network energy by the set of functions
$f$ for each base pair $(i,j)\in R_\ell$ of each input structure
$R_\ell$ that evaluate to $w_\ell \Ebp{k}{i,j,\val(x_i),\val(x_j)}$
under assignment $\val$.

@@

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, \emph{i.e.} such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp. Mat.~\ref{sec:weight_derivatives}). More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}. 
In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(e^\beta,\dots,e^\beta)$ and, at each iteration, generates a sample $\mathcal{S}$ of sequences. The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$. 
$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$. In practice, the constant $\gamma>1$ is chosen empirically to achieve effective optimization. 
While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior (choosing $\gamma=1.2$). Now, consider Given an elected vertex $v$ of a connected component, say the first node encountered in a rooted

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rationale for a rejection approach follows from: 
i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}). 
%Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances. 
For base pair-based energy models and special compatibility graph
(paths, cycles\ldots) this property rigorously follows from analytic
combinatorics, see \citet{Bender1983} and
\citet{Drmota1997}. In such cases, the expected number of
rejections before reaching the targeted energies remains constant when
$\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
$\varepsilon=0$. The \GCb-content of designs can also be controlled,
jointly and in a similar way, as done in
\Software{IncaRNAtion}~\citep{Reinharz2013}.

@@

Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps. At each step, we resampled (uniformly at random) the positions of a randomly selected component in the compatibility graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
%
Once again, for all instances, we observe a positive improvement of the quality of Boltzmann designs over uniform ones, suggesting a superior potential for optimization ($\sim 32\%$ avg improvement, Table~\ref{tab:benchmark-results}, optimized). This improvement is however more modest than in the absence of optimization, but we attribute this relative decay to the uniform nature of the resampling strategy, which somehow ends up canceling the benefits of the initial Boltzmann distribution, by uniformizing the sequence over longer runs. This suggests an alternative resampling strategy based on Boltzmann-distributed steps.

@@

\section{Counting valid designs is \#{\sf P}-hard}\label{sec:counting}

%%% reformulate later
%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given compatibility graph $G=(V,E)$. Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$. As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its compatibility graph is bipartite, which can be tested in linear time.
Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides. The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.

  Let us now hypothesize the existence of a polynomial-time algorithm $\mathcal{A}$ for \NumDesign over strongly-connected graphs $G$. Consider the (polynomial-time) algorithm $\mathcal{A}'$ that first executes $\mathcal{A}$ on $G$ to produce $\NumDesign(G)$, and returns $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$. Clearly $\mathcal{A}'$ solves $\#{\sf BIS}$ in polynomial-time. This means that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
\end{proof}

@@

\newpage
%\newgeometry{margin=2cm}
\onecolumn

@@

In addition, we generated four sets of instances of increasing
complexity. The instances of the sets RF3, RF4, RF5, and RF6, each
respectively specify 3,4,5, and 6 target structures for sequence
length 100.  For each instance (100 instances per set), we generated a
set of $k$ ($k=3,\dots,6$) compatible structures as follows
\begin{itemize}
\item Generate a random sequence of length 100;
\item Compute its minimum free energy structure (ViennaRNA package);
\item Add the new structure to the instances if the resulting compatibility graph is bipartite;
\item Repeat until $k$ structures are collected.
\end{itemize}
For each instance, we generated the dependency graphs in the base pair
model and in the stacking model. Then, we performed tree decomposition
(using strategy ``GreedyFillIn'' of LibTW~\citep{Dijk2006}) on each dependency
graph. The obtained treewidths are reported in
Fig.~\ref{fig:td-widths}, while Fig.~\ref{fig:td-times} shows the
corresponding run-times of the tree decomposition.
Fig.~\ref{fig:td-example} shows the tree decompositions for an example
instance from set RF3.

@@

\section{Exploiting the bipartite nature of compatibility graphs to reduce the complexity}\label{sec:improvedComplexity}

As already noted by \citet{Flamm2001}, the
requirement of canonical base pairing induces a bi-partition of each connected
component within the compatibility graph, such that the nucleotides assigned to the two set of nodes in the partition are 
restricted to values $\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively. It follows that, within each connected component of the compatibility graph, the assignment of any position drastically restricts the domain of the remaining positions, \emph{i.e.} they may now be assigned to only 2 out of the original 4 nucleotides.
\newcommand{\Parity}[1]{p_{#1}}
To exploit this property, and further reduce the exponential factor of the complexity, we introduce the notion of \Def{parity} of a vertex/position, whereby vertices of \Def{even} (resp. \Def{odd}) polarities may only be assigned values in $\{\Ab,\Gb\}$ (resp. $\{\Cb,\Ub\}$). Neighbors of an odd vertex must be even and vice versa, so the choice of a parity for any single vertex caracterizes the parity of all vertices in a bipartite connected subgraph.

Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the compatibility graph, we consider a modified version of the cluster tree, obtained by:
\begin{enumerate}
\item Electing, in each connected component $cc_i$, the first position $v_i^\star$ from the component encountered from the root in a pre-order traversal of the cluster tree;\label{step:election}
\item Introducing new variables $\mathcal{P}=\{\Parity{1},\ldots,\Parity{|CC|}\}$, representing the parity of elected vertices;
\item Adding, to each node $x$ in the tree decomposition, the parity variables of connected components that are represented in the node, \emph{i.e.} such that at least one of the positions in the connected component is in $\vars(x)$; \label{step:addparities}
\item Introducing, in the tree decomposition, a chain of intermediate nodes to fix cases where more than a single (parity) variable is introduced between a parent and its child.\label{step:intermediates}
\end{enumerate}
Note that, since connected components correspond to connected subtrees in the tree decomposition, then adding those variables preserves the properties of a tree decomposition. Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}. Finally, we note thatthe parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the compatibility graph. Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd. It follows that, thanks to the new variables introduced in Step~\ref{step:addparities}, the domain of each of the original variables can be reduced to two alternatives instead of the original four.

The computation of Alg.~\ref{alg:pf} on the new cluster tree, considering reduced assignments for position variables and additional odd/even assignments for parity variables, has complexity in $\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $n$ is the RNA length, $k$ is the number of structures, $w$ is the tree-width of the unmodified cluster tree, and $c$ is the maximum number of connected components represented in a node. It should be noted that $c\le \min(|CC|,w+1)$, so that the new complexity cannot be larger than the one stated in Th.~\ref{th:complexities}. Also, note that the number of nodes introduced by Step~\ref{step:intermediates} is bounded by $|CC|$, so that the introduction of these intermediate nodes does not impact the worst-case complexity. For the same reason, the complexity of the stochastic backtrack remains unchanged.
